---
title: "lab_02"
author: "derek willis"
date: "2023-09-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

Let's keep working with the UMD courses and Maryland campaign expense data we used in the pre_lab.

```{r}
umd_courses <- read_rds("data/umd_courses.rds")
maryland_expenses <- read_csv("data/maryland_expenses.csv")
```

## Answer questions

**Q1.** How many departments have offered the courses with "Climate" in the title and at least one seat? Which department has offered the most? 

**A1.** 26 departments have offered courses with "Climate" in the title with at at least one seat. The Atmospheric and Oceanic Science department offered the most courses. 

```{r}
Climate_courses_seats_1 <-
umd_courses |>
  filter(str_detect(title, "Climate")) |> filter(seats >= 1) |>
group_by(department) |>
  summarize(
    count_classes = n())|>
  arrange(desc(count_classes))
```

**Q2.** Of the departments listed in A2, which one is the most interesting to you, and why?

**A2.** It surprises me that the Public Health Science department only offers one course with the term Climate in it because I was a public health major in undergrad at UMD and I could have sworn we were required to take multiple climate courses.


**Q3** Which departments have the most courses in the Fall 2023 term? And which departments in that term have at least one syllabus posted? How are the top five departments different in each result?

**A3.** The top five departments with the most courses in the Fall 2023 term are business management, school of music, english, teaching and learning, policy and leadership, and information studies. Meanwhile, the departments in that term that have at least one syllabus posted include business management, economics, school of music, biological sciences programs, and information studies.

Business Management, Music, and Information Studies overlap here. These different functions, however, show that while the English Department and the Teach and Learning, Policy and Leadership Department offered a great deal of courses this term, they are severly lacking in updating their syllubi in comparison to their counterparts.

```{r}
umd_courses |> 
  filter(term == 202308)|> filter(syllabus_count >= 1) |>
  group_by(department) |>
    summarize(count_syllabus = n()) |>
    arrange(desc(count_syllabus))
```

**Q4.** Using the Maryland campaign expenses data, change the datatype of the expenditure_date column so that it is a date and not a chr, saving the result as the same dataframe. Then create a new column for the month based on expenditure_date. Then write code that shows total amount spent each month by Wes Moore's committee (you'll need to search the data to find the exact name). Write a sentence describing the pattern of spending over time.

**A4.** When we began collecting this data in May 2021, Wes Moore spending was a total of $6000.3. It then drastically increased the following month, almost tripling, and continued slightly growing (with obvious fluctuations) until it peaked in June and July of 2022. Aside from a random spike in October 2022, it has decreased significantly since it reached that aformentioned peak.

```{r}

maryland_expenses <- read_csv("data/maryland_expenses.csv")

maryland_expenses1 <- maryland_expenses |> 
  filter(str_detect(committee_name, "Moore  Wes For Maryland")) |>
  mutate(expenditure_date=mdy(expenditure_date)) |>
  mutate(month = floor_date(expenditure_date, "month")) |>
  group_by(month) |>
  summarise(total_amount = sum(amount))
```

**Q5.** Using case_when(), create a column in the Maryland expenses data called `spending_location` indicating whether each record spent money in Maryland or outside Maryland. For records that are in Maryland, make the new column's value "In-state" and for the others, make it "Out of state". Then write code that shows the total amount of money spent in each category and describe the results. You can do this in one statement or break it up into two statements (saving the new column to your dataframe).

**A5** This is super weird because it makes it seem like Wes Moore spent no money outside of Maryland? I find this really hard to believe, so I am worried that I did this incorrectly. He allegedly spent $77,723,146 in Maryland though, according to this data.

```{r}
maryland_expenses <- read_csv("data/maryland_expenses.csv")

maryland_expenses_with_state3 <- maryland_expenses |>
  mutate(
     spending_location = case_when(
        str_detect(address, " Maryland ") ~ "In-State",
        .default = "Out of State"
      )) |>
  group_by(spending_location) |>
  summarize(total = sum(amount))

```


**Q6.** Choose your own adventure: using either the UMD courses or campaign expenses data, ask and answer a question that is different than the previous ones. You can use one of the ones you provided in the pre_lab.

**A6.** I was going to be answering the following question: "For the UMD course data, which department offered the most Laws and Ethics courses in 2022?" but I keep getting this error and I'm not sure what it means: not supported in RHS call of a pipe.

```{r}
Laws_and_Ethics_Courses <-
umd_courses |>
  filter(str_detect(title, "Law")) |> filter(str_detect(title, "Ethics")) |> (filter(term > 202200)) |>
  group_by(department) |>
  summarize(
    count_classes = n()|>
  arrange(desc(count_classes)))
```
