---
title: "lab_08"
author: "derek willis"
date: "2023-04-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab. We'll be making some charts, working with dates and retrieving Census data.

```{r}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(janitor)
library(tidycensus)
```

## Load and modify data

**Task** Create a codeblock and load the following data from the data folder:

* Prince George's County 911 Overdose Calls

You will need to create columns for the date, week and month based on the existing `datetime` column.

```{r}
pg_overdoses <- read_csv("data/prince_georges_2022_overdoses.csv")

#Create columns for the month based on the existing `datetime` column

pg_calls_by_date <- pg_overdoses |>
  mutate(date = date(datetime)) |>
  mutate(month = month(datetime, label=TRUE)) |>
  mutate(week = week(datetime))

```

## Questions 

**Q1.** Which month saw the greatest percentage of total calls? Create a dataframe that calculates the percentage of all calls that each month's calls represents. Make a bar chart from that dataframe. Your bar chart must have:

* A clear title that states the main idea/finding
* Good labels for the x & y axis and a caption for the source, which is Prince George's County EMS.
* Readable bars - the values shouldn't be overlapping

Compare the results here to those from the pre_lab_08 bar chart - are there major differences in the months with the highest and lowest figures? Describe that below.

**A1.** The month with the greatest percentage of calls is December. This is drastically different from the Baltimore County data in the pre-lab, which had the warmest months with the most calls, specifically August. Baltimore County and PG County both had a cooler month lower in terms of percentage of calls, however, Baltimore County had February as the month with the least calls while PG County had January.

```{r}

pg_calls_by_date |>
  group_by(month) |> 
  summarize(total_calls = n()) |>
  mutate(
    total_calls_percent = (total_calls/1397) * 100) |>
  arrange(desc(total_calls)) |>
  ggplot() +
  geom_bar(aes(x=reorder(month,total_calls), weight=total_calls)) +
  coord_flip() + 
  theme_minimal() +
  labs(
    title="911 Calls Peak in Holiday Months",
    x = "month",
    y = "percent of total calls",
    caption = "Source: Prince George's County Emergency Medical Services"
  )

```

**Q2.** Let's visualize this data on a weekly basis using a line chart. As in Q1, generate a dataframe with the total number of calls for each week, and then create a line chart to show the distribution of calls over time. Your line chart must have:

* A clear title that states the main idea/finding
* Good labels for the x & y axis and a caption for the source, which is Prince George's County EMS.
* Readable labels

Describe the pattern of results; you may want to look at the data to dig into particular outliers.

**A2.** There is quite a bit of variation here, but most weeks stay within the 15% and 30% range with a distinct peak in what can be assumed to be in the very middle of the year (possibly week 27 or 28). Another peak is at the very end of the year. A few otehr smaller peaks can be seen as we approach the end of the year. These can both be considered outliers since they don't fall in the typical range.


```{r}
pg_calls_by_date |>
  group_by(week) |> 
  summarize(total_calls = n()) |>
  mutate(
    total_calls_percent = (total_calls/1397) * 100) |>
  
  ggplot() + 
  geom_line(aes(x=week, y=total_calls)) +
  labs(
    title="911 Calls Peak in Holiday Months",
    x = "week",
    y = "percent of total calls",
    caption = "Source: Prince George's County Emergency Medical Services"
  )
```

**Q3.**  A scatterplot is a type of chart that helps us see relationships between two variables. One variable goes on the x axis, the other on the y axis.  For each row/observation in our data, a scatterplot puts a circle (or a "point") where the two variables intersect on a grid. 

Statisticians use scatterplots to show graphically whether one variable is correlated -- related, in a statistical sense -- with another variable.  A classic example is the [relationship between ice cream sales and temperature](https://www.mathsisfun.com/data/scatter-xy-plots.html). The scatterplot below -- press play to load the image -- shows that relationship, that an increase in temperature is associated with an increase in ice cream sales. When it's 12C, sales are 200 dollars, and when it's hotter, 25C, sales are 600 dollars.

```{r}
knitr::include_graphics("https://www.mathsisfun.com/data/images/scatter-ice-cream1.svg")
```

We're going to use a scatterplot a little differently, to get a visual sense of two key variables: 

Our question is: does the median income in a zip code have any relationship to the number of overdose 911 calls in that zip code?

To answer this question, do the following:

1. Generate a dataframe with the number of 911 calls for each zip code.
2. Get data from the Census Bureau showing median household income for Maryland zip codes.
3. Join those two dataframes on their zip code columns, starting with the 911 calls dataframe.
4. Make a scatterplot showing the total calls and median income. I didn't show you how to do this, so look it up! Googling "ggplot scatterplot" is a good start.
5. Give it an appropriate title, source, and x and y axis titles.
6. Add a label for each point that shows the zip code using geom_text() - see some examples of its use at https://ggplot2.tidyverse.org/reference/geom_text.html#ref-examples. Try to make the names as easy to read as possible by avoiding overlap.
7. In the answer space below, describe what you see and answer the questions posed above. In a general sense, what do you think this means? Feel free to consider the actual raw values: how would you report out the main point(s) of this chart?

**A3.**  This is really ugly to me buuut I think it communicated exactly what was needed to be communciated. 

From this scatterplot, I am gathering that the zip codes with the greatest number of 911 overdose calls in PG county come from households in low socioeconomic class or households where the annual median income is around $75,000. 

```{r}
#Generate a dataframe with the number of 911 calls for each zip code.

pg_calls_by_zip <- pg_calls_by_date |>
  group_by(zipcode) |> 
  summarize(total_calls = n())

# Get data from the Census Bureau showing median household income for Maryland zip codes.

census_api_key("ccc0a99ac5ec5f6d3c490a4a8d38de4d555134ad", overwrite=TRUE)

acs5 <- load_variables(2021, "acs5", cache = TRUE)

median_income <- get_acs(geography = "zcta",
              variables = c(medianincome = "B19013_001"),
              year = 2021)

# Join those two dataframes on their zip code columns, starting with the 911 calls dataframe.

pgzip_with_acs <- pg_calls_by_zip |> inner_join(median_income, join_by("zipcode"=="GEOID"))

# Make a scatterplot showing the total calls and median income.

pgzip_with_acs |>
  ggplot(aes(x=total_calls, y=estimate)) + 
  geom_point() +
  labs(
    title="Most PG 911 Overdose Calls Come From Low-Income Areas",
    x = "Total Calls",
    y ="Estimated Household Income",
    caption = "Source: Prince George's County Emergency Medical Services, Census Data") +
  geom_text(aes(x=total_calls,
                y=estimate-5000,
                label=zipcode),
            check_overlap = TRUE)
```

