---
title: "lab_03"
author: "Derek Willis"
date: "2023-09-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse

# Load janitor
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab: the Maryland payments to vendors and Maryland grants and loans and, using janitor, change the columns so that they are lower-cased and any spaces are replaced with underscores.

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows
payments <- read_csv("data/State_of_Maryland_Payments_Data__FY2008_to_FY2024.csv", guess_max=373564)

# cleaning function
cleaned_payments <- payments |>
  clean_names()

# display the cleaned dataset
cleaned_payments

# Load in the Maryland grants and loan data
md_grants_loans <- read_csv("data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv")

# cleaning function
cleaned_md_grants_loans <- md_grants_loans |>
  clean_names()

# display the cleaned dataset
cleaned_md_grants_loans
```

## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed. Write code to find rows where the vendor name contains the county's school system, starting with a broad search and then narrowing it down. When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find? Look at & search the records in the dataframe while you construct your code.

A1: I found 87 records.

```{r}
# cleaning function
cleaned_payments |>
  filter(
    str_detect(vendor_name, 'PRINCE GEORGES COUNTY PUBLIC SCHOOL') | 
    str_detect(vendor_name, 'PRINCE GEORGES COUNTY SCHOOL') |
    str_detect(vendor_name,'PG COUNTY PUBLIC SCHOOL') |
    str_detect(vendor_name,'PRINCE GEORGES PUBLIC') |
    str_detect(vendor_name,'BD OF EDUCATION OF PG CO') |
    str_detect(vendor_name,'BOARD OF EDUCATION OF PRINCE GEORGE')) |>
  rename(pg_schools = vendor_name)

```

Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

A2: The fiscal year with the largest total was 2021 with a total of $147,0133,533. The aspect of the result that surprised me the most is the fact that it got rid of every other column. Maybe that's a little silly because these are the functions I used, but it worked out too quickly so I am a little skeptical. Another aspect that stands out to me is that the data for 2011, 2012, 2016, 2017, 2018, 2019, and 2020 are missing. Does this mean that no payments were made in that time? I find that hard to believe.

FIXED THE MISSING YEARS!!!

```{r}
# cleaning function
cleaned_payments |>
filter(
    str_detect(vendor_name, 'PRINCE GEORGES COUNTY PUBLIC SCHOOL') | 
    str_detect(vendor_name, 'PRINCE GEORGES COUNTY SCHOOL') |
    str_detect(vendor_name,'PG COUNTY PUBLIC SCHOOL') |
    str_detect(vendor_name,'PRINCE GEORGES PUBLIC') |
    str_detect(vendor_name,'BD OF EDUCATION OF PG CO') |
    str_detect(vendor_name,'BOARD OF EDUCATION OF PRINCE GEORGE')) |>
  rename(pg_schools = vendor_name) |>
  group_by(fiscal_year) |>
  summarise(
    total=sum(amount)
  ) |>
  
  arrange(desc(total))
```

Q3. You're interested in examining the vendors paid for capital projects at the University of Maryland. Search the payments data to find those payments, then write code to produce a list of each vendor and the total amount it received. Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? Provide at least three examples of vendor names that are similar and should be combined into one. Describe how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

A3. This data needs to be cleaned up to ensure that duplicates don't exist -- this can be determined not only by examining the vendor names in each row, but also the dates and zip codes from which these funds were received. Many vendors appear to be duplicates, including but not limited to the following: 
 - Whiting Turner Contracting, which is listed as Whiting Turner Contracting Co and Whiting Turner Contracting Company
 - Ayers Saint Gross, which is listed as Ayers Saint Gross Inc and Ayers Saint Gross Incorportated
 - Ballinger, which is listed as Ballinger Co and just Ballinger
I would clean this data by checking for duplicate rows using get_dupes(), which would identify all my duplicates in alphabetical order. 

```{r}
# cleaning function
cleaned_payments |>
  filter(str_detect(agency_name, 'UNIVERSITY OF MARYLAND')) |>
  group_by(vendor_name) |>
  summarise(
    total=sum(amount)
  ) |>
  
  arrange(desc(total))
```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. Once you find which columns contain the information you want, your code should standardize the data needed to find just the grants for stem cell research in a single filter. Then create a total amount of money for each fiscal year. Write a sentence summarizing how stem cell research funding has changed over time.

A4: I kept this dataset organized in order of fiscal year because I wanted to see the pattern, but in other circumstances I'd definitly want to see it in decreasing order. We collected this data from 2009 to 2020. We had a really large amount in 2009, in which more than $14 million came in. However, that amount has declined drastically (by double practically) and has mostly stayed at that level, never reaching that total since. 

```{r}
# cleaning function
cleaned_md_grants_loans |>
  filter(str_detect(description, 'Stem Cell') | str_detect(description, 'STEM CELL')) |>
  rename(stem_cell_projects = description) |>
  group_by(fiscal_year) |>
  summarise(
    total=sum(amount)
  )
```

Q5: Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

A5: According to this data, there are 18 grantees from this zip code (20742). This data definitely needs to be cleaned up though because even looking on the first page, I see bunch of duplicates for the general University of Maryland, which is listed under names like UNIVERSITY OF MARYLAND, University of Maryland College Park, and UNIVERSITY OF MD,COLLEGE PARK, as well as the UMD Research Administration and Advancement. Much like in A3 though, these duplicates are all listed under different variations of the same name, so I would need to identify these dupilcates and merge them in the dataset, which would unsurprisngly leave me with significantly less grantees than 18. The grantee most interesting to me is the UMD Data Center because while $159,441 is a substantial amount, I figured there would be signifcantly more money funneled here, especially when comparing it to the Clarice's funding.

```{r}
# cleaning function
cleaned_md_grants_loans |>
  filter(str_detect(zip_code, '20742')) |>
  group_by(grantee) |>
  summarise(
    total=sum(amount)
  ) |>
  arrange(desc(total))
```

