---
title: "week7_recap"
author: "Menna Ibrahim"
date: "2023-10-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## Loading the packages

Run the codeblock below to load the packages we will need for this recap

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
```

## Load Data

Run the codeblock below to load the data.

```{r}
earthquakes <- read_csv('https://earthquake.usgs.gov/earthquakes/feed/v1.0/summary/all_month.csv')

#Setting time column as datetime
earthquakes <- earthquakes |> mutate(time = as_datetime(time))
```

#### Answer the questions below

Most questions have a code block and a space for an answer below. Write the code you think is necessary and, in the answer space, write out what you did and what was the result.

------------------------------------------------------------------------

#### **Q1** Look at the earthquakes dataset. Finish the sentence below as if you were explaining the data to someone who had not seen it before but needs to know about it.

**A1:** This dataset contains information for all the earthquakes that have occurred in the past month. Information provided includes the time these earthquakes occurred, latitude and longitude as well as the place it occurred, the specific kind of event, such as earthquakes, quarry blast, and explosions, the magnitude, status, and more.

------------------------------------------------------------------------

#### **Q2** How many records there are there in this dataset? What do they mean and what useful information we can gather from it, looking at the columns?

**A2:** This datatset contains the records for the 9,774 earthquakes that have occurred in the past month since 9/11. The most important information it provides are the time these events occurred, the latitude and longitude as well as the place they occurred, the specific kinds of event (such as earthquakes, quarry blast, and explosions), and the magnitude.

------------------------------------------------------------------------

#### **Q3** How do I reorganize this data to see the ones that are the deepest first? What is the depth that shows up for the deepest one, and its magnitude?

```{r}
earthquakes |> 
  arrange(desc(depth))
```

**A3:** The depth for the deepest earthquake here, which occurred 3:35PM on 09/18/2023, is 669.9820. Its magnitude was 4.20.

------------------------------------------------------------------------

#### **Q4** I just want to see the earthquakes with a magnitude larger than 6. How do I do that? And how many are there that fit this criteria?

```{r}
earthquakes |> 
  filter(mag > 6)
```

**A4:** There are 13 earthquakes that fit this criteria.

------------------------------------------------------------------------

#### **Q5** What about if I want to see earthquakes that have both a magnitude larger than 6 and a depth smaller than 20? How many are there in the data set that fit [both]{.underline} these criteria?

```{r}
earthquakes |>
  filter(depth < 20 & mag > 6)
```

**A5:** There are 6 earthquakes that match this criteria.

------------------------------------------------------------------------

#### **Q6** What about if I want to see earthquakes that either have a magnitude larger than 6 OR a depth smaller than 20? How many are there in the data set that fit [either]{.underline} these criteria?

```{r}
earthquakes |>
  filter(depth < 20 | mag > 6)
```

**A6:** There are 7,446 earthquakes that match this criteria.

------------------------------------------------------------------------

#### **Q7** I'm interested in finding earthquakes that took place in Alaska. Which column should I look at? How do I use it to find all the earthquakes in Alaska? How many earthquakes took place there?

```{r}
earthquakes |>
  filter(str_detect(place, "Alaska"))
```

**A7:** There are 3,446 earthquakes that match this criteria

------------------------------------------------------------------------

#### **Q8** I notice that there is a column called 'type', that seems to have different kinds of tremors. What is the best way to find what are all the possible types of tremors, and counting how many of each there are in this data set? What are the first two most common types of tremors in this data set?

```{r}
earthquakes |>
  group_by(type) |>
  summarise(
    count=n()
  )|>
  arrange(desc(count))
```

**A8:** The two most common types are earthquakes and quarry blasts.

------------------------------------------------------------------------

#### **Q9** What is the average depth of the earthquake type in this data set? Is there anything that seems unusual or surprising in this finding?

```{r}
earthquakes |>
  filter(str_detect(type, "earthquake")) |>
  summarize(mean_depth = mean(depth))
```

**A9:** I think it's interesting that it doesn't even seem that deep. I don't earthquakes though, but I accidentally filtered for all the other types as well, and I thought it was crazy how much deeper the earthquake type is in comparison to the other types.

------------------------------------------------------------------------

#### **Q10** I'm interested, in the future, to see the hours in which earthquakes happen. How can I extract the hour from the time column?

```{r}
earthquakes |>
  mutate(hour = hour(time))
```

**A10:** See code above please!

------------------------------------------------------------------------

#### **Q11** I want to make a note of all the records in this data set that I consider serious. For my purposes, I'm thinking that all tremors that have a magnitude that is larger than 3 are serious. How do I automatically create a new column showing whether an earthquake is serious or not?

```{r}

earthquake_mag <- earthquakes |>
  mutate(magnitude_concern = case_when(
    (mag >3) ~ 'serious',
    .default = 'not serious')
  )
  
```

**A11:**Please see the code above!

------------------------------------------------------------------------

#### **Q12** I have no idea how earthquakes work and I'm interested in seeing if there is a particular time of day in which serious earthquakes happen. How can I see that condensed in a table with all the hours in a day and all the serious earthquakes in each hour? What is the hour with fewer serious earthquakes and the one with the most serious earthquakes?

```{r}
earthquake_mag |>
  mutate(hour = hour(time)) |>
  filter(str_detect(magnitude_concern, "serious")) |>
  group_by(hour) |>
  summarise(
    count=n()
  )|>
  arrange(desc(count))
```

**A12**11:00 AM is the hour with the most serious earthquakes, while 7:00 PM is the hour with the east serious earthquakes

------------------------------------------------------------------------

#### **Q13** What's another question you are interested in and how would you ask it in R?

Because we now know that 11:00AM is the the time most common for serious earthquakes, I am curious to know where those are located, so I can avoid being there before noon.

```{r}
earthquake_mag |>
  mutate(hour = hour(time)) |>
  filter(hour==11) |>
  group_by(place) |>
  summarise(
    count=n()
  )|>
  arrange(desc(count))

```
