---
title: "lab_01"
author: "Derek Willis"
date: "2023-01-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**). When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

Remember to add any new verbs or concepts to your class reference notebook!

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
```

## Load Data

We're going to load the UMD courses data.

**Task** Run the codeblock below to load the data, which is contained in an RDS file.

```{r}
umd_courses <- read_rds("lab_01.rds")
```

## Answer questions

**Q1** Which department has the highest mean (average) number of seats for its classes? You need to show the department with the highest average first. Which one is the most surprising/interesting to you?

**A1** The department with the highest mean number of seats is Engineering, Cooperative Education with an average number of 306.67 seats. The one most surprising to me is Air Science, as it has a relatively high number of average seats, however, I haven't met anyone in that department.

```{r}
umd_courses |>
  group_by(department) |>
  summarize(
    mean_seats = mean(seats)
  )|>
  arrange(desc(mean_seats))
```

**Q2** Find out the greatest number of times any individual class has been offered in our dataset. Your code must show the greatest number of classes first. 

**A2** The greatest number of times a course has been offered in our dataset is 72 times. This course is Directed Graduate Studies in Studio Art in the Studio Art department.

```{r}
umd_courses |>
  group_by(department, title) |>
  summarize(
    count_classes = n()
  )|>
  arrange(desc(count_classes))
```

**Q3** Count the number of courses for each instructor and department. Which instructor has taught the most? 

**A3** The instructor that has taught the most courses, according to this dataset, is Keith Yearwood in the Geographical Sciences department. This instructor has taught a total of 61 courses since we began collecting this data.

```{r}
umd_courses |>
  group_by(instructors, department)|>
  summarize(
    count_classes = n()
  )|>
  arrange(desc(count_classes))
```

**Q4** Given the results for Q3, what would have made it easier to find the answer? You don't have to know *how* to accomplish this; I want you to describe in words what would make the answer easier to see. How confident are you that the answer you got is the right one, and why? 

**A4** I had to click all the way to "page" 8 or rows 71-80 before I was able to find an instructor listed for a course, so it would have made this significantly easier had we omitted courses without an instructor listed. I am pretty confident in my result because I clicked previous each time, but I am weary that I missed someone along the way because I was bogged down by the N/As and TBDs listed beside each course department with a high number of courses listed. This was particularly problematic when I was just clicking pages hoping to find an instructor listed anywhere in the dataset.

**Q5** Take one of the questions you listed in the final question of the pre_lab and try to answer it here in code. If you don't know how to fully answer it, write what code you can and describe in words what else you would need to do. If you got an answer, is it interesting? 

**A5** I was trying to answer the following question: "Which department has grown the most since we have started collecting this data?" I didn't exactly know how to figure that out, so this is as far as I could get. This seems very inefficient though because it requires a significant amount of math and comparing each department with not only itself but other departments by the user. I initially tried to find the range using the range() function, but I couldn't figure out what to put in the () to get the departments AND terms that I wanted. I hope to figure this out soon because I would still need to manually subtract each 201712 term from each 202308 term for each department then compare them with each other find out which department exhibited the most growth.

```{r}
umd_courses |>
  group_by(department, term)|>
  summarize(
    maximum_seats = max(seats)
  )|>
arrange(desc(department))
```
