---
title: "lab_04"
author: "Derek Willis"
date: "2023-10-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

You'll need to load three packages for this: the tidyverse, lubridate and janitor.

**Task** load these three packages.

```{r}
# Turn off scientific notation
options(scipen=999)
options(scipen=999)
library(tidyverse)
library(lubridate)
library(janitor)

# load your packages

```

For this lab, we want to investigate spending by Maryland state and local candidates during the previous election. For example, we could ask how much money have candidates and committees spent on advertising on Facebook? Which vendors received the most money for media expenses? We have the data, but it's messy - names of payees are spelled differently and capitalized differently - and we need to clean it up. We'll use a combination of RStudio and OpenRefine to do that.

The basic process here is to start in RStudio, export data so that we can use OpenRefine to clean it up, and then bring it back into RStudio to finish our analysis.

## Load Data

You'll need to load one data set, a CSV file of Maryland campaign expenditures located in the data folder called "maryland_expenses.csv"

**Task** Create a codeblock below, then read the data in and assign it to an appropriate variable name. You'll want to clean up the column names and make sure that any date columns are actually date datatypes.

```{r}
# Load the Maryland expenditures data table
maryland_expenses <- read_csv("data/maryland_expenses.csv") 

# cleaning function
cleaned_maryland_expenses <- maryland_expenses |>
  clean_names() |>
  mutate(expenditure_date=mdy(expenditure_date))
```

## Answer questions

**Q1.** You've been assigned to write about Maryland political committees' use of Facebook advertising. There's a category column in the data that has values for different kinds of spending. One of them is "Media". You'll need to complete the following steps:

   1. Create a dataframe of Media expenses where the name of the payee is not NA.
   2. Then, using `write_csv`, make a CSV file of your media expenses in your data folder. `write_csv()` is similar to `read_csv()` but you provide it with the dataframe you're writing and the file path.
   3. Using OpenRefine, create a project using that CSV file, create a copy of the `payee_name` column called `payee_clean` and then standardize it, focusing on the payees that appear most often. Pay particular attention to variations of Facebook, bearing in mind that it changed its name in 2021: <https://about.fb.com/news/2021/10/facebook-company-is-now-meta/>), but also clean up variations of the most common payees.
   4. When you are done standardizing `payee_clean`, export it as a new CSV file to your data folder, *giving it a different name* and read it back into RStudio as a new dataframe.
   5. Using that `payee_clean` column in your new dataframe, calculate the total amount of Facebook spending by each committee and present the results showing the committee that spent the most first.

**A1.** The committees that spent the most are these in the following order: Friends of Jain Ashwani, Franchot Anderson-Walker for Maryland Slate, Friends Of Franchot Peter, Casa in Action PAC, and  Friends Of Billy Boniface

```{r}
media_expenses <- cleaned_maryland_expenses |>
  filter(str_detect(expense_category, 'Media')) |>
  drop_na(payee_name) |>
  arrange(expense_category)

write_csv(media_expenses, "data/media_expenses.csv")

facebook_expenses <- read_csv("data/media_expenses_facebook.csv")

facebook_expenses |> 
  filter(str_detect(payee_clean, 'Facebook')) |>
  group_by(committee_name) |>
  summarise(
    total=sum(amount)) |>
  arrange(committee_name) |>
  arrange(desc(total))
```

**Q2.** Let's do some research on the top committees from question 1. Look up the top 10 spenders: if they are candidate committees, what offices were they running for? If they were not candidates, what issues do they seem to care about? Write some sentences answering those questions, then go the Facebook ad library ([https://www.facebook.com/ads/library/?active_status=all&ad_type=political_and_issue_ads&country=US&sort_data\\[direction\\]=desc&sort_data\\[mode\\]=relevancy_monthly_grouped&media_type=all](https://www.facebook.com/ads/library/?active_status=all&ad_type=political_and_issue_ads&country=US&sort_data\%5Bdirection\%5D=desc&sort_data\%5Bmode\%5D=relevancy_monthly_grouped&media_type=all){.uri}) and look for ads from the top spender. What do you notice about the the information provided about those ads?

**A2.** 

PART A - 
1. Friends of Jain Ashwani: Arachana Gupta for Treasurer
2. Franchot Anderson-Walker for Maryland Slate: Monique Anderson Walker for PG councilwoman
3. Friends Of Franchot Peter: Peter Franchot for Governor
4. Casa in Action PAC: immigrant advocacy organization, often contibute to federal candidates
5. Friends Of Billy Boniface: Billy Boniface for Harford County Executive
6. Baron Jon (Jonathan) for Maryland: Jon Baron for Governor
7. Friends of Salimah Jasani: Salima Jasani for Baltimore Board of Education
8. WFP National Pac: advocates for the working contribution party, which is a left-wing minor political party in the US that
9. King and Siri for Maryland Slate: Michelle Siri, women's rights advocate, and John King for Governor
10. Friends of Eve Shuman: Eve Shuman for District 3, PG councilwoman

PART B -
1. When I looked up Friends of Ashwani Jain, I learned that the ads were launches in 2022. and htey are primarily inactive or have been removed off meta products like Facebook and Instagram. Furthermore, the majority of these ads seem to have lasted for one to five days, on average, and made between 5,000 and 20,00 impressions. That's a pretty big range, but I am noticing that the ads that the committee spent more money on are gaining more traction than their counterparts.

**Q3.** Given the list of top spenders in A1 and the primary results for governor (<https://elections.maryland.gov/elections/2022/primary_results/gen_results_2022_1.html>), who is missing from that Q1 list? Did those campaigns run Facebook ads? Who are the biggest recipients of media expenses by the Democratic primary winner for governor (HINT: that person has two committees, the second one with his running mate)?

**A3.** 

PART A -
Wes Moore, the current governor of Maryland, isn't showing up in this dataframe, which is confusing given that he won the whole election! Wes Moore absolutely ran some ads, as we can see here: https://www.facebook.com/ads/library/?active_status=all&ad_type=all&country=US&q=%22wes%20moore%22&sort_data[direction]=desc&sort_data[mode]=relevancy_monthly_grouped&search_type=keyword_exact_phrase&media_type=al, but it wasn't nearly as much as what we looked at before. Similarly, Dan Cox is not showing up, which again is strange because he is the winning republican candidate in this race.

PART B -
The biggest recipient is Moore Miller for Maryland Slate with 6,856,600.00 combined, which makes a lot of sense because that is Wes Moore and his running mate Aruna Miller. Other big recipients include Gusto, Paragon, and Sapphire Strategies. I did them each individually as well and found that they practically have all the same recipients, which is unsurprising since they ran together.

```{r}
cleaned_maryland_expenses |>
filter(
    str_detect(committee_name, 'Moore  Wes') |
    str_detect(committee_name, 'Miller  Aruna')) |>
  group_by(payee_name) |>
  summarise(
    total=sum(amount)) |>
  arrange(payee_name) |>
  arrange(desc(total))
```

**Q4.** The media expense data you cleaned also has a standardized purpose column. Write code to generate a list of the committees with the total amount of spending on online advertising. How does this list compare to the one from Q1 about top Facebook spenders? Of the top 5 committees who spent the most on Facebook ads from Q1, are there any that appear to have spent *all* of their online advertising dollars on Facebook?

**A4.** A few names in this top 5 list are identical to the one from Q1, including Casa in Action PAC. This committee seems to have spent about half its online advertising dollars on facebook. New names on this list though include Team Calvin Ball, MSEA's Fund for Children and Public Education, Maryland House Democratic Caucus Committee, and Baron-Williams for Maryland.

```{r}
media_expenses |>
  filter(
    str_detect(expense_purpose, 'Online Advertising')) |>
   group_by(committee_name, expense_purpose) |>
  summarise(
    total=sum(amount)) |>
  arrange(desc(total))
```

**Q5** Your editor asks you to summarize your findings about spending on Facebook by Maryland political candidates. How much do we know and what don't we know? What's the best story idea you've found here?

**A5** From this data, I've learned that while many candidates use Facebook as a marketing tool, the majority of the really successful ones put their efforts (and money) elsewhere. They are also very close to their running mate in terms of expenditures. We don't know much about payees listed as N/A, so I'm curious to know why recipients would choose to omit their information in that way. I am most curious to know about the folks that chose to use things other than Facebook and why that proved to be so successful when Facebook has been a longstanding reliable source for outreach and advertisement.  
